### 2022-06-01 LaTeX/report.tex
Updates

In the algorithm \ref{alg:seqsetbased}, the SCCs are progressively tracked in \texttt{sccs} and the equivalence relation $\mathcal{S}$ is updated with the \textsc{Unite} function. Note that the function written in Isabelle is different from the \textsc{Unite} function introduced earlier. From now on, \textsc{Unite} designates the following function, which was first written as a recursive function as follows:

@@

\isabelle{
    \bblue{definition} unite :: "\generic{v} $\Rightarrow$ \generic{v} $\Rightarrow$ \generic{v} env $\Rightarrow$ \generic{v} env" \where{}\\
    $~~~$"\blue{unite} \green{v w e} $\equiv$\\
    $~~~~~~$\bblue{let} \green{pfx} = takeWhile ($\lambda$\green{x}. \green{w} $\notin$ $\mathcal{S}$ \green{e x}) (stack \green{e});\\
    $~~~~~~~~~~$\green{sfx} = dropWhile ($\lambda$\green{x}. \green{w} $\notin$ $\mathcal{S}$ \green{e x}) (stack \green{e});\\
    $~~~~~~~~~~$\green{cc} = $\bigcup$ \{$\mathcal{S}$ \green{e x} $|$ \green{x}. \green{x} $\in$ set \green{pfx} $\cup$ \{hd \green{sfx}\}\}\\
    $~~~~~~$\bblue{in}  \green{e}\env{$\mathcal{S}$ := $\lambda$\green{x}. \bblue{if} \green{x} $\in$ \green{cc} \bblue{then} \green{cc} \bblue{else} $\mathcal{S}$ \green{e x}, stack := \green{sfx}}"
}

@@

In particular, the last conjunct expresses the fact that the equivalence relation on the stack is monotonic.

@@

Let us give some explanations:
\begin{itemize}
    \item[$\bullet$] The environment must be well-formed.
    \item[$\bullet$] The sub-graph induced by the nodes reachable from the node on which the function is called must be totally visited.
    \item[$\bullet$] The previous environment must be a sub-environment of the new one (\textit{i.e.} there is a monotonic ordering on the environments).
    \item[$\bullet$] The condition of reachability on the stack from the pre-condition must remain satisfied.
    \item[$\bullet$] The new stack is a suffix of the previous one (\textit{i.e.} it expressively represents the structure of FIFO stack).
    \item[$\bullet$] This one is a little more difficult to understand : it expresses the fact that if two nodes $m$ and $n$ such that $m \preceq n$ in the stack before the call on \texttt{dfs}, and if for all node $u$ on the stack, $u \Rightarrow^* v$ and $v \Rightarrow^* n$, then the execution of \texttt{dfs} has updated $\mathcal{S}$ such that $\mathcal{S}(m) = \mathcal{S}(n)$ in the environment returned by the function. This should be equated with the fourth conjunct : $v$ is reachable from all nodes on stack, and the fact that partials SCCs have a unique representative on stack.
    \item[$\bullet$] The last one is easier to understand: either $v$ is explored and the stack considered before and after the execution of the function has not changed, or the head of the stack is the representative of $v$ and $v$ had a representative in the previous stack.
\end{itemize}

\subsubsection{\texttt{dfss} pre- and post-conditions}
The pre-conditions on \texttt{dfss} are also rather simple, except the sixth conjunct which expresses the fact that before the execution of the function, if a node $n$ on the stack is reachable from $v$, then either $v \in \mathcal{S}(n)$, or $n$ is reachable from a successor of $v$. Once again, this condition has to be equated with the fact that all nodes on stack can reach $v$ (fifth conjunct in the following definition).

\isabelle{
\bblue{definition} pre\_dfss \where{}\\
"pre\_dfss \green{v vs e} $\equiv$\\
$~~~~$wf\_env \green{e}\\
$~~\land$ \green{v} $\in$ visited \green{e}\\
$~~\land$ \green{vs} $\subseteq$ successors \green{v}\\
$~~\land$ ($\forall$ \green{n} $\in$ set (stack \green{e}). reachable \green{n v})\\
$~~\land$ (stack \green{e} $\neq$ [])\\
$~~\land$ ($\forall$ \green{n}$\in$set(stack \green{e}).reachable \green{v n} $\longrightarrow$ \green{v}$\in$$\mathcal{S}$ \green{e n} $\lor$ ($\exists$ \green{m} $\in$ \green{vs}.reachable \green{m n}))\\
$~~\land$ (\green{v} $\in$ $\mathcal{S}$ \green{e} (hd (stack \green{e})))"
}

@@

The post-conditions on \texttt{dfss} are less complicated than the post-conditions of \texttt{dfs} and can easily be understood with the explanations of the other invariants.

@@

\subsubsection{\texttt{pre\_dfss} implies \texttt{pre\_dfs}}
This lemma fixes a node $w$ and assumes the pre-conditions on \texttt{dfss} on a node $v$, a set (of successors) $vs$ and an environment $e$. It assumes that $w$ is a successor of $v$ that is not visited, and it shows that the pre-conditions on \texttt{dfs} on $w$ the same environment $e$ are satisfied. The proof is quite as all conditions of \texttt{pre\_dfs} can be deduced from the conditions of \texttt{pre\_dfss}.

@@

\subsubsection{\texttt{pre\_dfs} implies \texttt{post\_dfs}}\label{lemma:pre_post_dfs}

@@

With these assumptions:
\begin{itemize}
    \item $\texttt{dfs\_dfss\_dom} (\texttt{Inr}(v,vs,e))$
    \item $\texttt{pre\_dfss}~v~vs~e$
    \item $vs \neq \varnothing$ and \\$\forall w \in vs,  w \notin \texttt{explored } e~\land~w \notin \texttt{visited } e~\land~\texttt{pre\_dfs}~w~e \Longrightarrow \texttt{post\_dfs}~w~e~(dfs~w~e)$
    \item let $e' =
    \begin{dcases}
		e & w \in \texttt{explored}~e \\
		\texttt{dfs}~v~e & w \in \texttt{visited}~e \\
        \texttt{unite}~v~w~e & \text{otherwise}
	\end{dcases}$\\
    $vs \neq \varnothing$ and \\$\forall w \in vs, \texttt{pre\_dfss}~v~(vs - \{w\})~e' \Longrightarrow \texttt{post\_dfss}~v~(vs - \{w\})~e'~(\texttt{dfss}~v~(vs~-~\{w\})~e')$
\end{itemize}
\noindent
We need to show the post-condition of \texttt{dfss}:
\begin{equation*}
    \texttt{post\_dfss}~v~vs~e~(\texttt{dfss}~v~vs~e)s
\end{equation*}

If the set of successors $vs$ is empty, then the post-condition of \texttt{dfss} is satisfied.
Otherwise, the proof is split according to the cases in the definition of $e'$. In the first case, $w$ is in the set of explored nodes of $e$. Hence, $e' = e$. The proof is rather straightforward thanks to the assumption on \texttt{pre\_dfss}. The following case, \textit{i.e.} $w \in \texttt{visited }e$ is much more complicated. In this case, $e' = \texttt{dfs}~v~e$.

@@

\end{document}

>>>>

In the algorithm \ref{alg:seqsetbased}, the SCCs are progressively tracked in \texttt{sccs} and the equivalence relation $\mathcal{S}$ is updated with the \textsc{Unite} function. Note that the function written in Isabelle is different from the \textsc{Unite} function introduced earlier. From now on, \texttt{unite} designates the following function, which was first written as a recursive function as follows:

@@

\BlankLine
\label{unite_def}
\isabelle{
    \bblue{definition} unite :: "\generic{v} $\Rightarrow$ \generic{v} $\Rightarrow$ \generic{v} env $\Rightarrow$ \generic{v} env" \where{}\\
    $~~~$"\blue{unite} \green{v w e} $\equiv$\\
    $~~~~~~$\bblue{let} \green{pfx} = takeWhile ($\lambda$\green{x}. \green{w} $\notin$ $\mathcal{S}$ \green{e x}) (stack \green{e});\\
    $~~~~~~~~~~$\green{sfx} = dropWhile ($\lambda$\green{x}. \green{w} $\notin$ $\mathcal{S}$ \green{e x}) (stack \green{e});\\
    $~~~~~~~~~~$\green{cc} = $\bigcup$ \{$\mathcal{S}$ \green{e x} $|$ \green{x}. \green{x} $\in$ set \green{pfx} $\cup$ \{hd \green{sfx}\}\}\\
    $~~~~~~$\bblue{in}  \green{e}\env{$\mathcal{S}$ := $\lambda$\green{x}. \bblue{if} \green{x} $\in$ \green{cc} \bblue{then} \green{cc} \bblue{else} $\mathcal{S}$ \green{e x}, stack := \green{sfx}}"
}

@@

In particular, the last conjunct expresses the fact that the equivalence relation on the stack is monotonic. This is a useful property as it explicitly gives a monotonic behaviour on the environment structure -- hence its name -- during the execution of the algorithm, even though we do not have an inductive rule on environments. Besides, this property is transitive, which will be used in the proofs.

@@

Let us give some explanations:
\begin{itemize}
    \item[$\bullet$] The environment must be well-formed.
    \item[$\bullet$] The sub-graph induced by the nodes reachable from the node on which the function is called must be totally visited.
    \item[$\bullet$] The previous environment must be a sub-environment of the new one (\textit{i.e.} there is a monotonic ordering on the environments).
    \item[$\bullet$] The condition of reachability on the stack from the pre-condition must remain satisfied.
    \item[$\bullet$] The new stack is a suffix of the previous one (\textit{i.e.} it expressively represents the structure of FIFO stack).
    \item[$\bullet$] The last one is easier to understand: either $v$ is explored and the stack considered before and after the execution of the function has not changed, or the head of the stack is the representative of $v$ and $v$ had a representative in the previous stack.
\end{itemize}

\subsubsection{\texttt{dfss} pre- and post-conditions}
The pre-conditions on \texttt{dfss} are also rather simple, except the sixth conjunct which expresses the fact that before the execution of the function, if a node $n$ on the stack is reachable from $v$, then either $v \in \mathcal{S}(n)$, or $n$ is reachable from a successor of $v$. Once again, this condition has to be equated with the fact that all nodes on stack can reach $v$ (fifth conjunct in the following definition). The definition is as follows:

\isabelle{
\bblue{definition} pre\_dfss \where{}\\
"pre\_dfss \green{v vs e} $\equiv$\\
$~~~~$wf\_env \green{e}\\
$~~\land$ \green{v} $\in$ visited \green{e}\\
$~~\land$ \green{vs} $\subseteq$ successors \green{v}\\
$~~\land$ ($\forall$ \green{n} $\in$ set (stack \green{e}). reachable \green{n v})\\
$~~\land$ (stack \green{e} $\neq$ [])\\
$~~\land$ (\green{v} $\in$ $\mathcal{S}$ \green{e} (hd (stack \green{e})))"
}

@@

The post-conditions on \texttt{dfss} are less complicated than the post-conditions of \texttt{dfs} and can easily be understood with the explanations of the other invariants:

@@

\subsubsection{\texttt{pre\_dfss} implies \texttt{pre\_dfs}}
This lemma fixes a node $w$ and assumes the pre-conditions on \texttt{dfss} on a node $v$, a set (of successors) $vs$ and an environment $e$. It assumes that $w$ is a successor of $v$ that is not visited, and it shows that the pre-conditions on \texttt{dfs} on $w$ the same environment $e$ are satisfied.

@@

The proof of the lemma is very simple since all conditions of \texttt{pre\_dfs} can be deduced from the conditions of \texttt{pre\_dfss}.

\subsubsection{\texttt{pre\_dfs} implies \texttt{post\_dfs}}\label{lemma:pre_post_dfs}
This lemma is probably the most important lemma -- and the most difficult to show -- in the proof of the algorithm. It assumes that the pre-condition on \texttt{dfs} are satisfied and shows that the post-condition on \texttt{dfs} are satisfied. It is stated as follows:

\isabelle{
\lm{} pre\_dfs\_implies\_post\_dfs:\\
$~~~$\fixes{} \blue{v e}\\
$~~~$\green{defines} "\blue{e1} $\equiv$ \blue{e}\env{visited := visited \blue{e} $\cup$ \{v\}, stack := (v \# stack \blue{e})}"\\
$~~~$\green{defines} "\blue{e'} $\equiv$ dfss \blue{v} (successors \blue{v}) \blue{e1}"\\
$~~~$\assumes{} 1: "pre\_dfs \blue{v e}"\\
$~~~~~~~$\and{} 2: "dfs\_dfss\_dom (Inl(\blue{v}, \blue{e}))"\\
$~~~~~~~$\and{} 3: "post\_dfss \blue{v} (successors \blue{v}) \blue{e1 e'}"\\
$~~~$\shows{} "post\_dfs \blue{v e} (dfs \blue{v e})"
}

Note that the definitions of $e_1$ and $e'$ follow from the definition of \texttt{dfs} which relies -- in an \textit{if} statement -- on the fact that $v = \text{hd } (\text{stack } e)$ or not, therefore the proof also considers this split of cases.

The first case assumes that $v = \text{hd}(\text{stack }e')$. Then with the second assumption ($v$ and $e$ are in the domain of definition of \texttt{dfs}), we can show that \texttt{dfs} returns the following environment:
\BlankLine
\noindent
$\texttt{dfs}~v~e = e'(\!|\texttt{sccs}:=\texttt{sccs}~e' \cup \{\mathcal{S}~e'~v\},\\
~~~~~~~~~~~~~~~~~~\texttt{explored}:=\texttt{explored}~e' \cup (\mathcal{S}~e'~v),\\
~~~~~~~~~~~~~~~~~~\texttt{stack}:=\texttt{tl}(\texttt{stack}~e')|\!)"$
\BlankLine
That is, it returns an environment $e_2 := \text{dfs}~v~e$ obtained from $e'$ in which a new SCC, namely $\mathcal{S}~e'~v$, is added. The different conjuncts in the definition of \texttt{post\_dfs} are then shown one by one. Some useful things to note -- and prove -- are the following:

\begin{itemize}
    \item $\text{stack }e_2 = \text{tl}(\text{stack }e')$, and since $v$ is the head of the stack of $e'$, we can show that $\text{stack }e_1 = \text{stack }e'$. The main idea is to use, from the assumption 3, the fact that:
    \begin{equation*}
        \exists~ns,~\text{stack } e_1 = ns~@~(\text{stack }e')
    \end{equation*}

    \item The equivalence relation is the same for $e'$ and $e_2$: $\forall x, \mathcal{S}~e'~x = \mathcal{S}~e_2~x$
    \item It is easy to show that $e$ is a sub-environment of $e_1$ (i.e. $\texttt{sub\_env}~e~e_1$). We can also show that $e_1$ is a sub-environment of $e_2$, which is much more difficult. However, the proofs in Isabelle are quite simple to read and understand. \textit{In fine}, by transitivity of the relation \texttt{sub\_env}, $e$ is a sub-environment of $e_2$.
    \item Showing that $\mathcal{S}~e'~v$ is a maximal SCC in the environment $e_2$ is a tricky part. The proof uses a contradiction-based reasoning and the fact that the equivalence relation is the same for $e'$ and $e_2$ to find a node being in a bigger SCC containing $\mathcal{S}~e'~v$ but not in $\mathcal{S}~e'~v$, and exploit the reachability from and to $v$ and the assumption 3.
\end{itemize}

In the second case, $v$ is not the head of the stack of $e'$, therefore $\texttt{dfs}~v~e = e'$ and the proof is much easier thanks to the post-condition on \texttt{dfss}, \textit{i.e.} assumption 3.

@@

With these assumptions:
\begin{itemize}
    \item $\texttt{dfs\_dfss\_dom} (\texttt{Inr}(v,vs,e))$
    \item $\texttt{pre\_dfss}~v~vs~e$
    \item $vs \neq \varnothing$ and \\$\forall w \in vs,  w \notin \texttt{explored } e~\land~w \notin \texttt{visited } e~\land~\texttt{pre\_dfs}~w~e \Longrightarrow \texttt{post\_dfs}~w~e~(\texttt{dfs}~w~e)$
    \item let $e' =
    \begin{dcases}
		e & w \in \texttt{explored}~e \\
		\texttt{dfs}~v~e & w \in \texttt{visited}~e \\
        \texttt{unite}~v~w~e & \text{otherwise}
	\end{dcases}$\\
    $vs \neq \varnothing$ and \\$\forall w \in vs, \texttt{pre\_dfss}~v~(vs - \{w\})~e' \Longrightarrow \texttt{post\_dfss}~v~(vs - \{w\})~e'~(\texttt{dfss}~v~(vs~-~\{w\})~e')$
\end{itemize}
\noindent
We need to show the post-condition of \texttt{dfss}:
\begin{equation*}
    \texttt{post\_dfss}~v~vs~e~(\texttt{dfss}~v~vs~e)s
\end{equation*}

If the set of successors $vs$ is empty, then the post-condition of \texttt{dfss} is satisfied.
Otherwise, the proof is split according to the cases in the definition of $e'$. In the first case, $w$ is in the set of explored nodes of $e$. Hence, $e' = e$. The proof is rather straightforward thanks to the assumption on \texttt{pre\_dfss}.

The following case, \textit{i.e.} $w \in \texttt{visited }e$ is much more complicated. In this case, $e' = \texttt{dfs}~v~e$. Altgouh the proof remains straightforward, the different properties are more difficult to establish. The main idea is to use the fact that $\texttt{post\_dfs}~w~e~e'$ is satisfied to show the pre-condition : $\texttt{pre\_dfss}~v~vs-\{w\}~e$. From the assumptions, we can obtain the post-condition : $\texttt{post\_dfss}~v~vs-\{w\}~e~e'$. There still remains some work to do inorder to show that this implies the final post-condition, \textit{i.e.} that $\texttt{post\_dfss}~v~vs-\{w\}~e~(\texttt{dfss}~v~vs-\{w\}~e)$ is satisfied.

In the last case, the function \texttt{unite} is called to compute the new environment $e'$. The proof uses the local variables defined in the definition of \texttt{unite}, \textit{i.e.} \texttt{pfx}, \texttt{sfx} and \texttt{cc} (see \ref{unite_def}). In particular, it is important to note that $\texttt{stack e} = \texttt{pfx } @ \texttt{ sfx}$ and $w$ is represented by a node in the stack being the head of \texttt{sfx}. A visual representation of the stack is given in \textsc{Figure \ref{fig:partial_correctness_unite}}. Morally, it is easy to understand that because of the cycle drawn on \textsc{Figure \ref{fig:partial_correctness_unite}}, $\mathcal{S}$ should be updated in the new environment so that all components of $\mathcal{S}$ for each node of \texttt{sfx} should be merged together.

\begin{figure}[!h]
    \ctikzfig{partial_correctness_unite}
    \caption{Representation of \texttt{stack e} in the case $w \in \texttt{ visited } e$. The black arrows represent reachability in both directions because the definition of a representative on the stack states that both nodes are in the same component of $\mathcal{S}$ for the environment $e$.}\label{fig:partial_correctness_unite}
\end{figure}

The approach is similar to the previous case : we prove the pre-condition on \texttt{dfss} and use the hypotheses to show the post-condition of \texttt{dfss}. The rest consists in proving the post-condition of \texttt{dfss} on $v$, $vs - \{w\}$ and the new environment $e'$ and the environment return by \texttt{dfss} implies the final post-condition.

@@

\end{document}x
